Carrier phase and amplitude estimation using pilots and data

### 2012-03-29 paper.tex
moved data so that it sits in it's own file
Added noncoherent plots.  ViterbiViterbi works very similarly to the Mackenthun algorithm.  It's not really clear how pilots should be included into ViterbiViterbi.  Need to run a simulations with a larger number of symbols to see if this biasing effect continues.
minor mods

The sequences in $S$ can be enumerated as follows.  Let $\sigma$ denote the permutation of the indicies in $D$ such that $z_{\sigma(i)}$ are in ascending order, i.e.
\begin{equation}\label{eq:sigmasortind}
z_{\sigma(i)} \leq z_{\sigma(k)}
\end{equation}
whenever $i < k $ where $i, k \in \{0, 1, \dots, \abs{D}-1\}$.  It is convenient to define the indices into $\sigma$ to be taken modulo $\abs{D}$, that is, if $m$ is an integer not from $\{0, 1, \dots, \abs{D}-1\}$ then we define $\sigma(m) = \sigma(k)$ where $k \equiv m \mod \abs{D}$ and $k \in  \{0, 1, \dots, \abs{D}-1\}$.  The first sequence in $S$ is 
\[
f_0 = f(0) = \{ b_i, i \in D \}
\]  
The next sequence $f_1$ is given by replacing the element $b_{\sigma(1)}$ in $f_0$ with $b_{\sigma(1)}e^{-j2\pi/M}$.  Given a sequence $x$ we use $x e_i$ to denote $x$ with the $i$th element replaced by $x_i e^{-j2\pi/M}$.  Using this notation,  
\[
f_1 = f_0 e_{\sigma(0)}.
\] 
The next sequence in $S$ is correspondingly 
\[
f_2 = f_0 e_{\sigma(0)} e_{\sigma(1)} = f_1 e_{\sigma(1)},
\]
and the $k$th sequence is
\begin{equation}\label{eq:fkrec}
f_{k+1} = f_{k} e_{\sigma(k)},
\end{equation}
In this way, all $(M+1)\abs{D}$ sequences in $S$ can be recursively computed.

We want to find the sequence $f_k \in S$ corresponding to the minimiser of~(\ref{eq:SSdatasymbols}).  A naive approach would be to compute $SS(f_k)$ for each sequence in $S$ and return the minimiser.  Computing $SS(f_k)$ for any particular $k$ requires $O(L)$ arithmetics operations.  So, this naive approach would require $O(L (M+1) \abs{D}) = O(L^2)$ operations in total.  Mackenthun~\cite{Mackenthun1994} showed how $SS(f_k)$ can be computed recursively.

Let,
\begin{equation}\label{eq:SSfk}
SS(f_k) = A - \frac{1}{L}\abs{Y_k}^2,
\end{equation}
where, 
\begin{align*}
Y_k = Y( f_k ) &= \sum_{i \in P} y_i p_i^*  + \sum_{i \in D} y_i f_{ki}^* \\
&= B + \sum_{i \in D}g_{ki},
\end{align*}
where $B = \sum_{i \in P} y_i p_i^*$ is a constant, independent of the data symbols, and $f_{ki}$ denotes the $i$th symbol in the sequence $f_k$, and for convenience, we put $g_{ki}  = y_i f_{ki}^*$.  Letting $g_{k} =\{ y_i f_{ki}^*, i \in D\}$ we have, from~\eqref{eq:fkrec}, that $g_k$ satisfies the recursive equation
\[
g_{k+1} = g_{k} e_{\sigma(k)}^*,
\]
where $g_{k} e_{\sigma(k)}^*$ indicates the sequence $g_k$ with the $\sigma(k)$th element replaced by $g_{k \sigma(k)}e^{j2\pi/M}$.  Now,
\[
Y_0 = B + \sum_{i \in D} g_{0i}
\] 
can be computed in $O(L)$ operations, and
\begin{align*}
Y_1 &= B + \sum_{i \in D} g_{1i} \\
&= B +  (e^{j2\pi/M} - 1)g_{0\sigma(0)} + \sum_{i \in D} g_{0i} \\
&= Y_0 + (e^{j2\pi/M} - 1) g_{0\sigma(0)},
\end{align*}
In general we have,
\[
Y_{k+1} = Y_k + (e^{j2\pi/M} - 1) g_{k\sigma(k)}.
\]
So, each $Y_k$ can be computed from it predecessor $Y_{k-1}$ in a constant number of operations.  Given $Y_k$, the value of $SS(f_k)$ can be computed in constant number of operations using~\eqref{eq:SSfk}.  Let $\hat{k}$ be integer the sequence that minimises $SS(f_k)$.  The least squares estimator of the complex amplitude is then computed according to~\eqref{eq:hata}, as 
\[
\hat{a} = \frac{1}{L} Y_{\hat{k}}.
\]
Psuedocode is given in Algorithm~\ref{alg:loglinear}.  This descibes a practical implementation of Mackenthun's algorithms.  Line~\ref{alg_return} contains the function $\operatorname{sortindicies}$ that, given the sequence $z = \{z_i, i \in D\}$, returns the permutation $\sigma$ of the indices in $D$ as desrcibed in~\eqref{eq:sigmasortind}.  Computing the $\operatorname{sortindicies}$ function requires sorting $\abs{D}$ elements.  This requires $O(L \log L)$ operations.  The $\operatorname{sortindicies}$ is the bottleneck in this algorithm.  All the other lines require only $O(L)$ operations to execute.  In the next sections we will show how to the sortinces function can be avoided.  This leads to an algorithm that requires only $O(L)$ operations.

\begin{algorithm} \label{alg:loglinear}
\SetAlCapFnt{\small}
\SetAlTitleFnt{}
\caption{Mackenthun's least squares estimator}
\dontprintsemicolon
\KwIn{$\{y_i, i \in P \cup D \}$}
\For{$i \in D$}{
$\phi = \angle{y_i}$ \;
$u = \frac{2\pi}{M}\round{\frac{M}{2\pi}\phi} $ \;
$z_i = \phi -  u $ \;
$g_i = y_i e^{-j u}$ \;
}
$\sigma = \operatorname{sortindicies}(z)$ \nllabel{alg_return} \;
$Y = \sum_{i \in P} y_i p_i^* + \sum_{i \in D} g_i $ \;
$\hat{a} = \frac{1}{L} Y$ \;
$\hat{Q} = \frac{1}{L}\abs{Y}^2$\;
\For{$k= 0$ \emph{\textbf{to}} $(M+1)\abs{D}-1$}{
$Y = Y + (e^{j2\pi/M} - 1) g_{\sigma(k)}$ \;
$g_{\sigma(k)} = g_{\sigma(k)} e^{j2\pi/M} $\;
$Q = \frac{1}{L}\abs{Y}^2$\;
\If{$Q > \hat{Q}$}{
 	$\hat{Q} = Q$ \;
 	$\hat{a} =  \frac{1}{L} Y$ \;
 }
}
\Return{$\hat{a}$ \nllabel{alg_return}}
\end{algorithm}

@@

\begin{lemma}
Let $\hat{a} = \hat{\rho}e^{j\hat{\theta}}$ so that $\hat{\theta}$ is the least squares estimator of the carrier phase and $\hat{\rho}$ is the least squares estimator of the amplitude.  Let 
\[
\lambda_i = \phi_i - \hat{\theta} - \frac{2\pi}{M}\round{\frac{M}{2\pi}(\phi_i - \hat{\theta})}.
\]
Then 
\[
\abs{\lambda_i} \leq \frac{1}{2M} - \frac{1}{2LM BLERG}
\]
for all $i \in P \cup D$.
\end{lemma}
\begin{proof}
\end{proof}

>>>>

The sequences in $S$ can be enumerated as follows.  Let $\sigma$ denote the permutation of the indicies in $D$ such that $z_{\sigma(i)}$ are in ascending order, that is,
\begin{equation}\label{eq:sigmasortind}
z_{\sigma(i)} \leq z_{\sigma(k)}
\end{equation}
whenever $i < k $ where $i, k \in \{0, 1, \dots, \abs{D}-1\}$.  It is convenient to define the indices into $\sigma$ to be taken modulo $\abs{D}$, that is, if $m$ is an integer not from $\{0, 1, \dots, \abs{D}-1\}$ then we define $\sigma(m) = \sigma(k)$ where $k \equiv m \mod \abs{D}$ and $k \in  \{0, 1, \dots, \abs{D}-1\}$.  The first sequence in $S$ is 
\[
f_0 = f(0) = \{ b_i, i \in D \}
\]  
The next sequence $f_1$ is given by replacing the element $b_{\sigma(1)}$ in $f_0$ with $b_{\sigma(1)}e^{-j2\pi/M}$.  Given a sequence $x$ we use $x e_i$ to denote $x$ with the $i$th element replaced by $x_i e^{-j2\pi/M}$.  Using this notation,  
\[
f_1 = f_0 e_{\sigma(0)}.
\] 
The next sequence in $S$ is correspondingly 
\[
f_2 = f_0 e_{\sigma(0)} e_{\sigma(1)} = f_1 e_{\sigma(1)},
\]
and the $k$th sequence is
\begin{equation}\label{eq:fkrec}
f_{k+1} = f_{k} e_{\sigma(k)},
\end{equation}
In this way, all $(M+1)\abs{D}$ sequences in $S$ can be recursively enumerated.

We want to find the sequence $f_k \in S$ corresponding to the minimiser of~(\ref{eq:SSdatasymbols}).  A naive approach would be to compute $SS(f_k)$ for each sequence in $S$ and return the minimiser.  Computing $SS(f_k)$ for any particular $k$ requires $O(L)$ arithmetic operations.  So, this naive approach would require $O(L (M+1) \abs{D}) = O(L^2)$ operations in total.  Mackenthun~\cite{Mackenthun1994} showed how $SS(f_k)$ can be computed recursively.

Let,
\begin{equation}\label{eq:SSfk}
SS(f_k) = A - \frac{1}{L}\abs{Y_k}^2,
\end{equation}
where, 
\begin{align*}
Y_k = Y( f_k ) &= \sum_{i \in P} y_i p_i^*  + \sum_{i \in D} y_i f_{ki}^* \\
&= B + \sum_{i \in D}g_{ki},
\end{align*}
where $B = \sum_{i \in P} y_i p_i^*$ is a constant, independent of the data symbols, and $f_{ki}$ denotes the $i$th symbol in the sequence $f_k$, and for convenience, we put $g_{ki}  = y_i f_{ki}^*$.  Letting $g_{k} =\{ y_i f_{ki}^*, i \in D\}$ we have, from~\eqref{eq:fkrec}, that $g_k$ satisfies the recursive equation
\[
g_{k+1} = g_{k} e_{\sigma(k)}^*,
\]
where $g_{k} e_{\sigma(k)}^*$ indicates the sequence $g_k$ with the $\sigma(k)$th element replaced by $g_{k \sigma(k)}e^{j2\pi/M}$.  Now,
\[
Y_0 = B + \sum_{i \in D} g_{0i}
\] 
can be computed in $O(L)$ operations, and
\begin{align*}
Y_1 &= B + \sum_{i \in D} g_{1i} \\
&= B +  (e^{j2\pi/M} - 1)g_{0\sigma(0)} + \sum_{i \in D} g_{0i} \\
&= Y_0 + \eta g_{0\sigma(0)},
\end{align*}
where $\eta = (e^{j2\pi/M} - 1)$.  In general we have,
\[
Y_{k+1} = Y_k + \eta g_{k\sigma(k)}.
\]
So, each $Y_k$ can be computed from it predecessor $Y_{k-1}$ in a constant number of operations.  Given $Y_k$, the value of $SS(f_k)$ can be computed in constant number of operations using~\eqref{eq:SSfk}.  Let $\hat{k}$ be integer the sequence that minimises $SS(f_k)$.  The least squares estimator of the complex amplitude is then computed according to~\eqref{eq:hata}, as 
\[
\hat{a} = \frac{1}{L} Y_{\hat{k}}.
\]
Psuedocode is given in Algorithm~\ref{alg:loglinear}.  This descibes a practical implementation of Mackenthun's algorithm.  Line~\ref{alg_sortindices} contains the function $\operatorname{sortindicies}$ that, given the sequence $z = \{z_i, i \in D\}$, returns the permutation $\sigma$ of the indices in $D$ as described in~\eqref{eq:sigmasortind}.  Computing the $\operatorname{sortindicies}$ function requires sorting $\abs{D}$ elements.  This requires $O(L \log L)$ operations.  The $\operatorname{sortindicies}$ function is the primary bottleneck in this algorithm.  The loops on lines~\ref{alg_loop_setup} and~\ref{alg_loop_search} and the operations on lines~\ref{alg_Y} to lines~\ref{alg_Q} all requirse $O(L)$ or less operations.  In the next sections we will show how to the sortinces function can be avoided.  This leads to an algorithm that requires only $O(L)$ operations.

\begin{algorithm} \label{alg:loglinear}
\SetAlCapFnt{\small}
\SetAlTitleFnt{}
\caption{Mackenthun's least squares estimator}
\dontprintsemicolon
\KwIn{$\{y_i, i \in P \cup D \}$}
\For{$i \in D$ \nllabel{alg_loop_setup}}{
$\phi = \angle{y_i}$ \;
$u = \frac{2\pi}{M}\round{\frac{M}{2\pi}\phi} $ \;
$z_i = \phi -  u $ \;
$g_i = y_i e^{-j u}$ \;
}
$\sigma = \operatorname{sortindicies}(z)$ \nllabel{alg_sortindices} \;
$Y = \sum_{i \in P} y_i p_i^* + \sum_{i \in D} g_i $ \nllabel{alg_Y}\;
$\hat{a} = \frac{1}{L} Y$ \;
$\hat{Q} = \frac{1}{L}\abs{Y}^2$ \nllabel{alg_Q} \;
\For{$k= 0$ \emph{\textbf{to}} $(M+1)\abs{D}-1$ \nllabel{alg_loop_search}}{
$Y = Y + (e^{j2\pi/M} - 1) g_{\sigma(k)}$ \;
$g_{\sigma(k)} = g_{\sigma(k)} e^{j2\pi/M} $\;
$Q = \frac{1}{L}\abs{Y}^2$\;
\If{$Q > \hat{Q}$}{
 	$\hat{Q} = Q$ \;
 	$\hat{a} =  \frac{1}{L} Y$ \;
 }
}
\Return{$\hat{a}$ \nllabel{alg_return}}
\end{algorithm}

@@

\begin{lemma}
Let $\hat{a} = \hat{\rho}e^{j\hat{\theta}}$ so that $\hat{\theta}$ is the least squares estimator of the carrier phase and $\hat{\rho}$ is the least squares estimator of the amplitude.  Let 
\[
\lambda_i = \phi_i - \hat{\theta} - \frac{2\pi}{M}\round{\frac{M}{2\pi}(\phi_i - \hat{\theta})}.
\]
Then 
\[
\abs{\lambda_i} \leq \frac{1}{2M} - \frac{1}{2LM BLERG}
\]
for all $i \in P \cup D$.
\end{lemma}
\begin{proof}
Recall that $\hat{d}_i = \hat{d}_i(\hat{\theta})$ are the minimsers of $SS$.  The proof now proceeds by contradiction.  Assume that $\lambda_i > \frac{1}{2M} - \frac{1}{2LM BLERG}$ for some $k$.  Put $r_i = \hat{d}_i e_k$, then
\begin{align*}
Y(\{ r_i, i \in D\}) &= \frac{1}{L}\left( B + \sum_{i \in D} y_i\hat{r}_i^* \right) \\
&= \hat{Y} + \frac{1}{L}\eta y_k\hat{d}_k^*.
\end{align*}
where we put $\hat{Y} = Y(\{ \hat{d}_i, i \in D\})$.  Now,
\begin{align*}
SS(\{r_i, i \in D\}) &= A - \frac{1}{L} \abs{ Y(\{ r_i, i \in D\}) }^2 \\
&= A - \frac{1}{L} \abs{ \hat{Y} + \frac{1}{L} \eta y_k\hat{d}_k^* }^2 \\
&= A - \frac{1}{L}\abs{\hat{Y}}^2 - \frac{2}{L^2}\Re\left(\eta \hat{Y}^* y_k \hat{d}_k^* \right) -  \frac{1}{L^3}\abs{ \eta y_k \hat{d}_k^* }^2\\
&= SS(\{\hat{d}_i, i \in D\}) - R
\end{align*}
where 
\[
R = \frac{2}{L^2}\Re\left(\eta \hat{Y}^* y_k \hat{d}_k^* \right) +  \frac{1}{L^3}\abs{ \eta y_k \hat{d}_k^* }^2.
\]
We will show that $R$ is positive, from which is will follow that $SS(\{r_i, i \in D\}) < SS(\{\hat{d}_i, i \in D\})$, violating the fact that $\{\hat{d}_i, i \in D\}$ are minimisers of $SS$.

To see that $R$ is positive write
\begin{align*}
R &\geq  \frac{2}{L^2}\abs{\hat{Y} y_k}\Re\left(\eta e^{j(\phi_i - \hat{\theta})}\hat{d}_k^* \right) +  \frac{1}{L^3}\abs{ \eta}^2 \abs{ y_k }^2 \\
&= \frac{1}{L^2}\abs{y_k} \left( 2\abs{\hat{Y}}\Re\left(\eta e^{j(\phi_i - \hat{\theta})} \hat{d}_k^* \right) + \frac{1}{L}\abs{ \eta}^2 \abs{ y_k }  \right) 
\end{align*}
so it is sufficient to show that
\[
\Re\left(\eta e^{j(\phi_i - \hat{\theta})} \hat{d}_k^* \right) > - \frac{\abs{ \eta}^2 \abs{ y_k }}{2 L \abs{\hat{Y}}}.
\]
Unfinished.  This proof will work, but the problem is that $\abs{y_k}$ can be really small.  The would appear to neccesitate a very large number of buckets.  In some sense, this is no problem, but you have to be able to allocate memory for all of these?  Atleast you need a way of ordering them, up to a certain resolution, without sorting.  I'm not convinced this will ever be possible.
\end{proof}
